gratia: An R package for exploring generalized additive models

### 2024-06-28 paper.Rmd
in hindsight, simplifying the code is best; added a note in the text now to say I'm also excluding terms in the other linear predictors.

Perhaps we are interested in the average expected chlorophyll *a* between 40--50 degrees N and 40--50 degrees W. It would be quite a simple matter to calculate this value from the fitted model: we first create a slice through the data for the spatial locations were are interested in using the `data_slice()` function, which ensures that `ds` contains everything we need to predict from the fitted model
```{r avg-chl-ds, dependson = "fit-chl-lsslmodel", cache = TRUE}
ds <- data_slice(m2,
  lat = evenly(lat, lower = 40, upper = 50, by = 0.5),
  lon = evenly(lon, lower = -50, upper = -40, by = 0.5))
```
Next, `fitted_values()` returns the predicted values at the specified locations. I only include the spatial effects, excluding the effects of ocean depth and day of year:
```{r avg-chl-fv, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}
use <- c("(Intercept)", "s(lat,lon)", "s.1(lat,lon)", "s.2(lat,lon)")
fv <- fitted_values(m2, data = ds, terms = use) # predict
```
Finally, I summarise the predictions for the location parameter to yield the average of the predicted values
```{r avg-chl-sumamrise, dependson = "avg-chl-fv", cache = TRUE}
fv |>
  filter(.parameter == "location") |>
  summarise(chl_a = mean(.fitted))
```
While this is an acceptable answer to the question, it lacks an uncertainty estimate. This is where posterior sampling is useful. With a small modification of the above code and a little data wrangling, we can produce an uncertainty estimate ,using `fitted_samples()` to generate posterior draws of the expected chlorophyll *a*:
```{r post-sampling-1, warning = FALSE, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}
fs <- fitted_samples(m2,     # model
  data = ds,                 # values of covariates to predict at
  terms = use,               # which terms to include in predictions
  n = 10000,                 # number of posterior draws
  method = "gaussian",       # Gaussian approximation to the posterior
  unconditional = TRUE,      # incl uncertainty for estimating lambda
  n_cores = 4,               # how many CPU cores to compute MVN samples
  seed = 342)                # set the random number seed, used internally
```
The posterior draws can then be summarised as before, except now the average chlorophyll *a* is calculated separately for each posterior draw (`.draw`)
```{r summarise-posterior, dependson = "post-sampling-1", cache = TRUE}
fs |>                                 # take the posterior draws
  group_by(.draw) |>                  # group them by `.draw`
  summarise(chl_a = mean(.fitted)) |> # compute mean of fitted chl a
  ggdist::median_qi()                 # summarise posterior
```
The posterior distribution of average chlorophyll *a* is summarised using `median_qi()` from the *ggdist* package [@Kay2024-rv; @Kay2024-uj]. While it would be a simple matter to compute the interval with base R commands, the use of `median_qi()` illustrates how *gratia* tries to interact with other packages.

>>>>

Perhaps we are interested in the average expected chlorophyll *a* between 40--50 degrees N and 40--50 degrees W. It would be quite a simple matter to calculate this value from the fitted model: we first create a slice through the data for the spatial locations were are interested in using the `data_slice()` function, which ensures that `ds` contains everything we need to predict from the fitted model
```{r avg-chl-ds, dependson = "fit-chl-lsslmodel", cache = TRUE}
ds <- data_slice(m2,
  lat = evenly(lat, lower = 40, upper = 50, by = 0.5),
  lon = evenly(lon, lower = -50, upper = -40, by = 0.5))
```
Next, `fitted_values()` returns the predicted values at the specified locations. I only include the spatial effects, excluding the effects of ocean depth and day of year, and ignore terms in the other linear predictors as they do not affect the expected chlorophyll *a*:
```{r avg-chl-fv, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}
use <- c("(Intercpt)", "s(lat,lon)")
fv <- fitted_values(m2, data = ds, terms = use) # predict
```
Finally, I summarise the predictions for the location parameter to yield the average of the predicted values
```{r avg-chl-sumamrise, dependson = "avg-chl-fv", cache = TRUE}
fv |>
  filter(.parameter == "location") |>
  summarise(chl_a = mean(.fitted))
```
While this is an acceptable answer to the question, it lacks an uncertainty estimate. This is where posterior sampling is useful. With a small modification of the above code and a little data wrangling, we can produce an uncertainty estimate ,using `fitted_samples()` to generate posterior draws of the expected chlorophyll *a*:
```{r post-sampling-1, warning = FALSE, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}
fs <- fitted_samples(m2,     # model
  data = ds,                 # values of covariates to predict at
  terms = use,               # which terms to include in predictions
  n = 10000,                 # number of posterior draws
  method = "gaussian",       # Gaussian approximation to the posterior
  unconditional = TRUE,      # incl uncertainty for estimating lambda
  n_cores = 4,               # how many CPU cores to compute MVN samples
  seed = 342)                # set the random number seed, used internally
```
The posterior draws can then be summarised as before, except now the average chlorophyll *a* is calculated separately for each posterior draw (`.draw`)
```{r summarise-posterior, dependson = "post-sampling-1", cache = TRUE}
fs |>                                 # take the posterior draws
  group_by(.draw) |>                  # group them by `.draw`
  summarise(chl_a = mean(.fitted)) |> # compute mean of fitted chl a
  ggdist::median_qi()                 # summarise posterior
```
The posterior distribution of average chlorophyll *a* is summarised using `median_qi()` from the *ggdist* package [@Kay2024-rv; @Kay2024-uj]. While it would be a simple matter to compute the interval with base R commands, the use of `median_qi()` illustrates how *gratia* tries to interact with other packages.
